---
title: "Mixed ANOVA"
author: "Alexandra Stephenson (a1795202)"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load}
pacman::p_load(tidyverse, readr, rstatix, ggpubr)
data <- read_csv(here::here("data/2023-03-01_gene-data.csv"))
plot_height <- 5
```

# Example
Source: https://www.datanovia.com/en/lessons/mixed-anova-in-r/

```{r example load}
library(datarium)
data("performance", package = "datarium")
performance <- performance %>%
  gather(key = 'time', value = 'score', t1, t2) %>%
  convert_as_factor(id, time)
```

## Visualisation
```{r example vis}
bxp <- ggboxplot(performance,
                 x = 'gender',
                 y = 'score',
                 color = 'stress',
                 palette = 'jco',
                 facet.by = 'time')
bxp
```

## Check assumptions
```{r example check outliers}
performance %>%
  group_by(gender, stress, time) %>%
  identify_outliers(score)
```

```{r example check normality}
normality_test <- performance %>%
  group_by(gender, stress, time) %>%
  shapiro_test(score)
normality_test %>% arrange(p)
```

```{r example check normality plot}
ggqqplot(performance, 'score', ggtheme = theme_bw()) +
  facet_grid(time ~ stress, labeller = "label_both")
```

```{r example check homogeneity}
homogeneity_test <- performance %>%
  group_by(time) %>%
  levene_test(score ~ gender*stress)
homogeneity_test %>% arrange(p)
```

## Three-way mixed ANOVA
```{r example ANOVA}
res.aov <- anova_test(data = performance,
                      dv = score,
                      wid = id,
                      within = time,
                      between = c(gender, stress))
get_anova_table(res.aov)
```

## Post-hoc
### Two-way interaction at each time level
```{r example post-hoc 1}
two.way <- performance %>%
  group_by(time) %>%
  anova_test(dv = score, wid = id, between = c(gender, stress))
two.way
```

### Simple simple main effect
```{r example post-hoc 2}
stress.effect <- performance %>%
  group_by(time, gender) %>%
  anova_test(dv = score, wid = id, between = stress)
stress.effect# %>% filter(time == "t2")
```

### Simple simple pairwise comparisons
```{r example post-hoc 3}
pwc <- performance %>%
  group_by(time, gender) %>%
  pairwise_t_test(score ~ stress, p.adjust.method = "bonferroni") %>%
  select(-p, -p.signif) # Remove details
pwc %>% filter(time == "t2", gender == "female")
```

### Visualization: box plots with p-values
```{r example viz results}
pwc <- pwc %>% add_xy_position(x = "gender")
pwc.filtered <- pwc %>% filter(time == "t2", gender == "female")
bxp +
  stat_pvalue_manual(pwc.filtered, tip.length = 0, hide.ns = TRUE) +
  labs(
    subtitle = get_test_label(res.aov, detailed = TRUE),
    caption = get_pwc_label(pwc)
  )
```

# Actual data
```{r data load}
data_aov <- data %>%
  mutate(CL = `cell line`) %>%
  select(-`cell line`) %>%
  gather(key = "conc", value = "GE", '0','1','2','3','4','5','6','7','8','9','10') %>%
  convert_as_factor(sheet_names) %>%
  filter(!is.na(GE))
head(data_aov)
```

## Visualisation
```{r data viz}
bxp <- ggdotplot(data_aov,
                 x = 'conc',
                 y = 'GE',
                 color = 'treatment',
                 palette = 'jco',
                 facet.by = 'CL',
                 binwidth = 1)
bxp
ggsave(here::here("figures/conc-ge-grouping_scatterplot.svg"), width = 2*plot_height, height = plot_height)
```

## Check assumptions
```{r data outliers}
data_aov %>%
  group_by(CL, treatment, conc) %>%
  identify_outliers(GE)
```
Therefore, no extreme outliers.

```{r data normality}
normality_test <- data_aov %>%
  group_by('CL', treatment, conc) %>%
  shapiro_test(GE)
normality_test %>% arrange(p)
```
Only one potentially non-normal.

```{r data normality plots}
ggqqplot(data_aov, 'GE', ggtheme = theme_bw()) +
  facet_grid(conc ~ treatment, labeller = "label_both")
```
None appear non-normal from this plot.

```{r data homogeneity, warning = FALSE}
homogeneity_test <- data_aov %>%
  group_by(conc) %>%
  levene_test(GE ~ CL*treatment)
homogeneity_test %>% arrange(p)
```
Non-homogenous (NOT SURE WHAT TO DO HERE?)

## Three-way mixed ANOVA
```{r data ANOVA}
res.aov <- anova_test(data = data_aov,
                      dv = GE,
                      wid = sheet_names,
                      within = conc,
                      between = c(CL, treatment))
get_anova_table(res.aov)
```
There is a statistically significant three-way interaction between cell line, treatment, and concentration.

## Post-hoc
### Two-way interaction at each cell line level
```{r data post-hoc 1, warning = FALSE}
two.way <- data_aov %>%
  group_by(CL) %>%
  anova_test(dv = GE, wid = sheet_names, between = c(conc, treatment))
two.way
```
There is a statistically significant interaction between concentration and treatment for cell line 'wild-type'.

### Simple simple main effect
```{r data post-hoc 2}
treatment.effect <- data_aov %>%
  group_by(CL, conc) %>%
  anova_test(dv = GE, wid = sheet_names, between = treatment)
treatment.effect %>% 
  data.frame() %>% 
  mutate(conc = as.integer(conc)) %>%
  arrange(conc) %>%
  arrange(desc(CL)) %>%
  arrange(desc(`p..05`)) %>% 
  filter(CL == "wild-type")
```
There is a statistically significant effect of treatment on all concentration levels for the 'wild-type' cell line.

### Simple simple pairwise comparisons
```{r data post-hoc 3}
pwc <- data_aov %>%
  filter(treatment == "placebo") %>% 
  group_by(treatment, CL) %>%
  pairwise_t_test(GE ~ conc, p.adjust.method = "bonferroni") %>%
  select(-p, -p.signif)
pwc %>% 
  filter(CL == "wild-type", p.adj.signif != 'ns') %>% 
  mutate(group1 = as.integer(group1), group2 = as.integer(group2)) %>% 
  arrange(p.adj.signif) %>% 
  arrange(group2) %>% 
  arrange(group1)
```
```{r}
pwc2 <- data_aov %>%  
  filter(treatment != "placebo", conc != 5) %>% 
  group_by(treatment, CL) %>% 
  pairwise_t_test(GE ~ conc, p.adjust.method = "bonferroni") %>%
  select(-p, -p.signif)
pwc2_print <- pwc2 %>% 
  filter(p.adj.signif != 'ns') %>% 
  mutate(group1 = as.integer(group1), group2 = as.integer(group2)) %>% 
  arrange(p.adj.signif) %>% 
  arrange(group2) %>% 
  arrange(group1)
pwc2_print
```
There are statistically significant different gene expressions between the following pairs of concentration:

Treatment | Group 1 | Group 2
----------|---------|---------
placebo   | 0       | 6
placebo   | 0       | 10
placebo   | 1       | 10
placebo   | 2       | 6
placebo   | 2       | 10
placebo   | 4       | 6
placebo   | 4       | 10
placebo   | 5       | 10
AF 42     | 0       | 4
AF 42     | 0       | 6
AF 42     | 0       | 7
AF 42     | 0       | 8
AF 42     | 0       | 9
AF 42     | 0       | 10
AF 42     | 1       | 4
AF 42     | 1       | 6
AF 42     | 1       | 7
AF 42     | 1       | 8
AF 42     | 1       | 9
AF 42     | 1       | 10
AF 42     | 2       | 6
AF 42     | 2       | 7
AF 42     | 2       | 8
AF 42     | 2       | 9
AF 42     | 2       | 10
AF 42     | 3       | 7
AF 42     | 3       | 8
AF 42     | 3       | 9
AF 42     | 3       | 10
AF 42     | 4       | 8
AF 42     | 4       | 9
AF 42     | 4       | 10
AF 42     | 6       | 10

**DO MORE HERE**

```{r data post-hoc 3}
pwc3 <- data_aov %>%
  filter(CL != "wild-type") %>%
  group_by(treatment, CL) %>%
  pairwise_t_test(GE ~ conc, p.adjust.method = "bonferroni") %>%
  select(-p, -p.signif) %>% 
  filter(p.adj.signif != 'ns') %>% 
  mutate(group1 = as.integer(group1), group2 = as.integer(group2)) %>% 
  arrange(p.adj.signif) %>% 
  arrange(group2) %>% 
  arrange(group1)
pwc3 %>% 
  filter(CL != "wild-type", p.adj.signif != 'ns') %>% 
  mutate(group1 = as.integer(group1), group2 = as.integer(group2)) %>% 
  arrange(p.adj.signif) %>% 
  arrange(group2) %>% 
  arrange(group1)
```

### Visualization: box plots with p-values
```{r data viz results 1}
pwc <- pwc %>%
  add_xy_position(x = "conc") %>%
  mutate(group1 = as.integer(group1),
         group2 = as.integer(group2)) %>%
  arrange(p.adj.signif)
head(pwc)
pwc2 <- pwc2 %>% 
  add_xy_position(x = "conc") %>%
  mutate(group1 = as.integer(group1),
         group2 = as.integer(group2)) %>%
  arrange(p.adj.signif)
head(pwc2)
```

```{r data viz results 2}
pwc.filtered <- pwc %>% filter(CL == "wild-type", treatment == "placebo")
head(pwc.filtered)
pwc2.filtered <- pwc2# %>% filter(CL == "wild-type")
head(pwc2.filtered)
```

```{r data viz results 3}
bxp2 <- ggdotplot(data_aov,
                 x = 'conc',
                 y = 'GE',
                 color = 'treatment',
                 palette = 'jco',
                 facet.by = 'CL',
                 binwidth = 1,
                 size = 2)
bxp2
```

```{r data viz results 4}
bxp2 +
  stat_pvalue_manual(pwc.filtered, tip.length = 0, hide.ns = TRUE) +
  stat_pvalue_manual(pwc2.filtered, tip.length = 0, hide.ns = TRUE) +
  labs(
    subtitle = get_test_label(res.aov, detailed = TRUE),
    caption = get_pwc_label(pwc)
  )
ggsave(here::here("figures/conc-ge-grouping_scatterplot_labelled.svg"), width = 2*plot_height, height = plot_height)
```

## Report

**DO MORE HERE**

A three-way mixed ANOVA was performed to evaluate the effects of treatment, concentration and cell line on gene expression.

There were no extreme outliers, as assessed by box plot method. The data was normally distributed, as assessed by Shapiro-Wilk’s test of normality ($p > 0.05$) and seen in the qqplot of the data. 
<!-- There was homogeneity of variances ($p > 0.05$) as assessed by Levene’s test of homogeneity of variances. -->

There was a statistically significant three-way interaction between treatment, concentration and cell line, $F(10, 30) = 2.244, p = 0.0043$.

For the simple two-way interactions and simple simple main effects, a Bonferroni adjustment was applied leading to statistical significance being accepted at the $p < 0.025$ level.

There was a statistically significant simple two-way interaction between concentration and treatment for cell line 'wild-type', $F(10, 21) = 26.1$, $p = 1.29\times 10^{-9}$, but not for cell line 'cell-type 101', $F(10,22) = 0.895$, $p = 0.553$.

There was a statistically significant simple simple main effect of treatment on gene expression for all concentration levels and the cell line 'wild-type', $F(1, 2) = 54.288$ $p = 0.018$, $F(1, 2) = 68.081$ $p = 0.014$, $F(1, 2) = 368.331$ $p = 0.003$, $F(1, 2) = 120.454$ $p = 0.008$, $F(1, 2) = 86.894$ $p = 0.011$, $F(1, 2) = 212.206$ $p = 0.044$, $F(1, 2) = 53.816$ $p = 0.018$, $F(1, 2) = 500.484$ $p = 0.002$, $F(1, 2) = 207.936$ $p = 0.005$, $F(1, 2) = 202.365$ $p = 0.005$ and $F(1, 2) = 108.000$ $p = 0.009$.

All simple simple pairwise comparisons were run between the different treatment groups for all concentration levels and cell line 'wild-type'. A Bonferroni adjustment was applied.

The mean performance score was statistically significantly different between 0 mg/ml and 6 mg/ml ($p = 0.0146$), 0 mg/ml vs 10 mg/ml ($p = 0.00331$), 1 mg/ml vs 10 mg/ml ($p = 0.0108$), 2 mg/ml vs 6 mg/ml ($p = 0.0224$), 2 mg/ml vs 10 mg/ml ($p = 0.00488$), 4 mg/ml vs 6 mg/ml ($p = 0.0394$), 4 mg/ml vs 10 mg/ml ($p = 0.00816$), and 5 mg/ml vs 10 mg/ml ($p = 0.0401$), all for cell line 'wild-type' and treatment 'placebo'. 
There was no significant difference between the other pairs of concentration levels.


