---
title: "Modelling the effect of growth factor, treatment and cell line on gene expression"
author: "Alexandra Stephenson"
date: today
date-format: long
format: 
  pdf:
    self-contained: true
    df-print: tibble
editor_options:
  chunk_output_type: console
execute:
  warning: false
  message: false
  echo: false
bibliography: references.bib
---

```{r}
pacman::p_load(tidyverse, readr, lme4, knitr, performance, sjPlot, lmerTest)
options(knitr.kable.NA = "")
theme_set(theme_light())
data <- read_csv("data/2023-03-01_gene-data.csv")
data_long <- data %>%
  mutate(CL = `cell line`, treat = treatment) %>%
  unite(`cell line`, `treat`, sep = "; ", col = "grouping") %>%
  pivot_longer(cols = 4:14, names_to = "concentration", values_to = "GE") %>%
  filter(GE >= 0) %>%
  mutate(concentration = as.integer(concentration),
         GL = as.factor(sheet_names),
         CL = as.factor(CL),
         treatment = as.factor(treatment),
         grouping = as.factor(grouping))
```

# Introduction
The data set consists of eight gene lines, each with one of two cell lines (wild-type or cell-type 101), one of two treatments (the placebo treatment or activating factor 42), and eleven different concentrations of growth factor (recorded in mg/ml). Thus, for each pair of cell line and treatment, there are two gene lines, each with eleven concentrations of growth factor (from $0$ mg/ml to $10$ mg/ml). 

This report investigates the effect of growth factor concentration, treatment, and cell line on gene expression, as well as the effect of gene line.

# Method

The data were cleaned and analysed using the R language [@r-2022], and the packages knitr [@knitr], lme4 [@lme4], lmerTest [@lmerTest], performance [@performance], readr [@readr], sjPlot [@sjPlot] and tidyverse [@tidyverse]. 
Any data points recorded as $-99$ were taken to indicate no data was recorded, or `NA`. Only one data point is missing, that of the gene expression for growth factor concentration $5$ mg/ml and gene line GL-fUg (with cell line wild-type and treatment activating factor 42). 
Data entries that referred to the same cell line or treatment were cleaned so that they all used the same exact phrase (for example, the abbreviation '`'WT' was replaced by 'wild-type').

Exploratory data analysis was then conducted on the data, including plotting gene expression versus cell line, gene expression versus treatment, and gene expression versus growth factor concentration.

Several fixed effects and mixed effects models were then fitted, using ANOVA tests to confirm the statistical significance of all terms in the models, and compared using Akaike's Information Criterion (AIC), $R^{2}$ values and root mean squared error (RMSE).

# Results

Initially, an investigation of the parameters that may impact gene expression was conducted prior to fitting any models.

```{r fig-ge-cl-boxplot, fig.cap = "A boxplot of gene expression for each cell line (wild-type and cell-type 101).", out.width="80%"}
data_long %>%
  ggplot(aes(x = CL, y = GE, col = CL)) +
  geom_boxplot() +
  theme(legend.position = 'none') +
  harrypotter::scale_color_hp_d("Ravenclaw") +
  labs(x = "Cell line",
       y = "Gene expression")
```

A boxplot of gene expression, grouped by cell line, is shown in @fig-ge-cl-boxplot. From this boxplot, it can be seen that there does not appear to be a significant difference between gene expression for wild-type and gene expression for cell-type 101. This suggests that cell line may not be a predictor of gene expression, or at least, not on its own.

```{r fig-ge-treat-boxplot, fig.cap = "A boxplot of gene expression for each treatment (placebo and activating factor 42).", out.width="80%"}
data_long %>%
  ggplot(aes(x = treatment, y = GE, col = treatment)) +
  geom_boxplot() +
  theme(legend.position = 'none') +
  harrypotter::scale_color_hp_d("Ravenclaw") +
  labs(x = "Treatment",
       y = "Gene expression")
```

@fig-ge-treat-boxplot shows a boxplot of gene expression for each treatment type (placebo or activating factor 42). From this boxplot, it can be seen that there does appear to be a difference between gene expression for placebo and gene expression for activating factor 42, and that activating factor 42 increases gene expression. This suggests that treatment is a predictor of gene expression.

```{r fig-ge-conc-grouping, fig.cap = "A plot of gene expression as a function of concentration, coloured by cell line (wild-type or cell-type 101) and treatment (placebo or activating factor 42).", out.width="80%"}
data_long %>%
  ggplot(aes(x = concentration, y = GE, color = grouping)) +
  geom_point() +
  ylim(0, NA) +
  harrypotter::scale_color_hp_d("Ravenclaw") +
  labs(x = "Concentration (mg/ml)",
       y = "Gene expression",
       color = "Cell Line; Treatment")
```

The effect of concentration of growth factor on gene expression is shown in @fig-ge-conc-grouping, with data points coloured according to cell line and treatment.
<!-- The data is plotted in @fig-ge-conc-grouping, with gene expression on the y axis and concentration on the x axis, with the data points coloured by cell line and treatment.  -->
There appears to be a relationship between concentration and gene expression, which suggests that concentration is a predictor of gene expression.
@fig-ge-conc-grouping also shows that there appear to be differences between the pairs (cell-type 101, placebo) and (wild-type, placebo) and the other two pairs of cell line and treatment. However, there does not appear to be a difference between the pair (cell-type 101, activating factor 42) and  the pair (wild-type, activating factor 42). This suggests that for the placebo treatment, cell line has an impact on gene expression, but for the activating factor, cell line may not have an impact on gene expression. Thus, cell line may be a predictor for gene expression, possibly in an interaction term.

<!-- **REPEATED MEASURES HERE** -->
The initial analysis suggests that gene expression may be predicted by treatment and concentration, with an interaction effect with cell line.
Given that the gene expression for each cell line and treatment was measured for different concentrations of growth factor for the same gene line, then this must also be considered when fitting models on the data.

## Fixed effects model

A fixed effects model can be fit using the `step` function to select the best model based on AIC, where the full scope is gene expression as a function of concentration, treatment, and cell line, with interaction terms between all three predictors. 
<!-- Gene line is not included as it is an additional parameter to be controlled for, not a possible predictor to be included in the model.  -->
Using AIC, the function selects the full model as the best model.
```{r}
lm_null <- lm(GE ~ 1, data = data_long)
scope <- GE ~ concentration*treatment*CL
lm_step <- step(lm_null, scope = scope, direction = "both", trace = 0)
```

The statistical significance of the terms in the selected fixed effects model can be found using an ANOVA test. @tbl-sjPlot-step shows the results of the ANOVA test conducted upon the selected fixed effects model, from which it can be seen that all of the terms are statistically significant.

```{r}
#| label: tbl-sjPlot-step
#| tbl-cap: "An ANOVA table showing the statistical significance of each predictor in the fixed effects model."
lm_step <- lm(GE ~ concentration*treatment*CL, data = data_long)
anova(lm_step) %>% kable(digits = c(0, 5, 5, 6, 6))
```

## Mixed effects models

The fixed effects model does not consider the impact of gene line. To consider this as a factor, a model can be fitted for gene expression as a function of concentration, with gene line as a random effect. The residuals plot for this model is shown in @fig-m1-plot. This figure shows that there is still variance not explained by concentration alone, so this model will not be considered further.

```{r}
m1 <- lmer(GE ~ concentration + (1|GL), data = data_long, na.action = na.omit)
```
```{r fig-m1-plot, fig.cap = "The residuals plot for the model of gene expression as a function of concentration, with gene line as a random effect.", out.width="80%"}
plot(m1)
```

The next model considered fits gene expression as a function of concentration and treatment (with interaction terms), as well as gene line as a random effect. The residuals plot for this model is shown in @fig-m2-plot, which shows that there does not appear to be any residual variance for this model.

```{r}
m2 <- lmer(GE ~ concentration*treatment + (1|GL), data = data_long, na.action = na.omit)
```
```{r fig-m2-plot, fig.cap = "The residuals plot for the model of gene expression as a function of concentration and treatment (with interaction terms), with gene line as a random effect.", out.width="80%"}
plot(m2)
```

Two models are fitted that include cell line as a predictor. One with interaction terms between concentration and treatment only, and one with interaction terms between concentration, treatment and cell line. The residuals plots for these models are shown in @fig-m3-plot and @fig-m4-plot, respectively. For both of these residuals plots, there does not appear to be any residual variance.

```{r}
m3 <- lmer(GE ~ concentration*treatment + CL + (1|GL), data = data_long, na.action = na.omit)
```
```{r fig-m3-plot, fig.cap = "The residuals plot for the model of gene expression as a function of concentration and treatment (with interaction terms) and cell line (without any interaction terms), with gene line as a random effect.", out.width="80%"}
plot(m3)
```
```{r}
m4 <- lmer(GE ~ concentration*treatment*CL + (1|GL), data = data_long, na.action = na.omit)
```
```{r fig-m4-plot, fig.cap = "The residuals plot for the model of gene expression as a function of concentration, treatment and cell line (with interaction terms between all three predictors), with gene line as a random effect.", out.width="80%"}
plot(m4)
```

<!-- The residuals plots in @fig-m2-plot, @fig-m3-plot and @fig-m4-plot all show that there does not seem to be any residual variance for these models. -->

@tbl-sjPlot-m3 shows the statistical significance of each term in the mixed effects model with concentration and treatment (with the interaction term) and cell line (without interactions with this predictor) as predictors, as well as the gene line random effect. From this table, it can be seen that the cell line predictor is not statistically significant, so this term should be removed from the model. Removing this term results in the mixed effects models with concentration and treatment predictors (with the interaction term) and the gene line random effect. The statistical significance of each term in this model is shown in @tbl-sjPlot-m2, from which it can be seen that all fixed effect terms in this model should be kept.

```{r}
#| label: tbl-sjPlot-m3
#| tbl-cap: "An ANOVA table showing the statistical significance of each fixed effect predictor in the mixed effects model with concentration, treatment and cell line predictors, and an interaction term between concentration and treatment, as well as a gene line random effect. The cell line term is not statistically significant, whilst the interaction term between concentration and treatment is, so the individual concentration and treatment fixed effect terms should also be retained, whilst the cell line term is removed."
anova(m3) %>% kable(digits = c(5, 5, 0, 6, 6, 6))
```

```{r}
#| label: tbl-sjPlot-m2
#| tbl-cap: "An ANOVA table showing the statistical significance of each fixed effect predictor in the mixed effects model with concentration and treatment predictors, and an interaction term, as well as a gene line random effect. The interaction term between concentration and treatment is significant, so all fixed effect terms in the model should be kept."
anova(m2) %>% kable(digits = c(5, 5, 0, 6, 6, 6))
```

@tbl-sjPlot-m4 shows the statistical significance of each term in the model with concentration, treatment and cell line as predictors, along with interaction terms between all predictors, and a gene line random effect. From this table, it can be seen that the interaction term between concentration, treatment and cell line is statistically significant, so this term should be kept. Because this term should be kept, all of the other fixed effect terms should also be retained. 

```{r}
#| label: tbl-sjPlot-m4
#| tbl-cap: "An ANOVA table showing the statistical significance of each fixed effect predictor in the mixed effects model with concentration, treatment and cell line predictors, and interaction terms between all of the predictors, as well as a gene line random effect. The three-way interaction term between concentration, treatment and cell line is statistically significant, so all fixed effect terms in the model should be retained."
anova(m4) %>% kable(digits = c(5, 5, 0, 6, 6, 6))
```

The statistical significance of the random effect terms in the mixed effects models can also be tested, as shown in @tbl-ranova-m2 and @tbl-ranova-m4. From these tables, it can be seen that the random effect term in each model is statistically significant, and should therefore be retained.

```{r}
#| label: tbl-ranova-m2
#| tbl-cap: "The statistical significance of the random effect term in the mixed effects model with concentration and treatment (and the interaction term between these predictors) as fixed effects, showing that the gene line random effect term is statistically significant."
ranova(m2) %>% kable(digits = c(0, 4, 4, 5, 4, 25))
```

```{r}
#| label: tbl-ranova-m4
#| tbl-cap: "The statistical significance of the random effect term in the mixed effects model with concentration, treatment and cell line (and the interaction terms between these predictors) as fixed effects, showing that the gene line random effect term is statistically significant."
ranova(m4) %>% kable(digits = c(0, 4, 4, 5, 4, 25))
```

## Comparing the models

```{r}
#| label: tbl-performance-m234
#| tbl-cap: "The AIC, $R^{2}$ values and root mean squared errors for each of the three fitted models."
fixed_effects_model <- lm_step
mixed_effects_model_1 <- m2
mixed_effects_model_2 <- m4
compare_performance(fixed_effects_model, mixed_effects_model_1, mixed_effects_model_2) %>% 
  select(c("Name", "AIC", "R2", "R2_conditional", "RMSE")) %>% 
  rename("conditional $R^{2}$" = "R2_conditional", "$R^{2}$" = "R2") %>% 
  kable(digits = c(0,4,7,7,6))
```

The fixed effects model and mixed effects models can be compared to each other using AIC values, $R^{2}$ values and RMSE values (shown in @tbl-performance-m234).
These show that the mixed effects model with interaction terms between concentration, treatment and cell line has the best AIC. The other mixed effects model, where cell line is not a predictor, has a very similar AIC value, whilst the fixed effects model (without the gene line random effect) has the worst AIC.
The conditional $R^{2}$ values for the mixed effects models, which take into account both the fixed effects and the random effects, are very similar for both models, but the model with all interaction terms is still slightly better. The $R^{2}$ value for the fixed effects model is worse than the conditional $R^{2}$ values for the mixed effects models.
The root mean squared error for the fixed effects model is much greater than the root mean squared error for either mixed effects model, whilst the mixed effects model that does not include cell line as a predictor has a larger root mean squared error than the mixed effects model that does include cell line as a predictor. This suggests that the mixed effects model with interaction terms between all three predictors is the best.
<!-- The root mean square errors of the mixed effects model that  cell line in @tbl-performance-m234 are very similar, whilst the root mean squared error for the fixed effects model is much greater than the other values. The lowest root mean squared error occurs for the mixed effects model with interaction terms between all three predictors, suggesting that this model is the best. -->
Thus, all three metrics indicate that the model with interaction terms between all three predictors, and with gene line as a random effect, is the best model.


# Discussion

```{r fig-ge-conc-grouping-model, fig.cap = "A plot of gene expression as a function of concentration, coloured by gene line (with cell line and treatment also indicated), and with the fitted model indicated by the lines."}
data_long %>%
  mutate(group = grouping, geneline = GL) %>%
  unite(group, geneline, sep = "; ", col = "grouping2") %>%
  ggplot(aes(x = concentration, y = GE, color = grouping2)) +
  geom_point() +
  geom_line(aes(y = predict(m4))) +
  ylim(0, NA) +
  harrypotter::scale_color_hp_d("Ravenclaw") +
  labs(x = "Concentration (mg/ml)",
       y = "Gene expression",
       color = "Cell Line; Treatment; Gene Line")
```

The chosen model is the mixed effects model with concentration, treatment and cell line as predictors, along with all interaction terms between the three predictors, and gene line as a random effect. This model is indicated by the lines in @fig-ge-conc-grouping-model, where each line is the fitted model for a different gene line. This figure shows how the gene lines with the placebo treatment (in darker brown and lighter blue) have a flatter slope than the gene lines with the activating factor 42 treatment (in lighter brown, grey and darker blue). The slope of the fitted model for the gene lines with the wild-type cell line and the placebo treatments (in darker brown) is also flatter than the slope of the fitted model for the gene lines with the cell-type 101 cell line and the placebo treatments (in lighter blue).

The coefficients of the fitted model are shown in @tbl-fixed-coefs-m4, and the random intercepts are shown in @tbl-random-coefs-m4. Because a random intercept term is included (for gene line), then the intercept for each gene line is found as the average intercept (in @tbl-fixed-coefs-m4) plus the gene line specific intercept in @tbl-random-coefs-m4.

```{r}
#| label: tbl-fixed-coefs-m4
#| tbl-cap: "The coefficients of the chosen model. The value of the intercept is the average intercept, which is added to the values in @tbl-random-coefs-m4 to find the intercept for each gene line."
fixef(m4) %>% 
  data.frame() %>% 
  rename(value = ".") %>%
  kable()
```

```{r}
#| label: tbl-random-coefs-m4
#| tbl-cap: "The difference from the average intercept (shown in @tbl-fixed-coefs-m4) for each gene line."
random_effects <- ranef(m4)$GL
random_effects %>%
  rename(value = `(Intercept)`) %>% 
  kable()
```

From these tables, it can be seen that as growth factor concentration increases, so does gene expression. It can also be seen that the placebo treatment has a smaller intercept and flatter slope than the activating factor 42 treatment does. Similarly, the wild-type cell line has a lower intercept and flatter slope than the cell-type 101 cell line does. 
Thus, gene expression is higher for higher concentrations of the growth factor, the activating factor 42 treatment and cell-type 101 cell line. Conversely, lower concentrations of the growth factor, the placebo treatment and wild-type cell line results in lower gene expression.

<!-- $$ -->
<!-- \text{gene expression} = 9.9175 + \text{gene line intercept} -4.215909(\text{placebo treatment}) -0.3615634(\text{wild-type cell line}) + 0.0817907(\text{placebo treatment})(\text{wild-type cell line}) + (3.0514091 -1.4055(\text{placebo treatment}) - 0.1214545(\text{wild-type cell line}) - 0.9674091(\text{placebo treatment})(\text{wild-type cell line}))(\text{concentration}) -->
<!-- $$ -->

<!-- The fitted model has the coefficients shown in @tbl-coefs-m4. -->

<!-- ```{r} -->
<!-- #| label: tbl-coefs-m4 -->
<!-- #| tbl-cap: "**CAPTION**" -->
<!-- fixed_effects <- fixef(m4) -->
<!-- random_effects <- ranef(m4)$GL -->
<!-- coefs <- random_effects -->
<!-- # orig_rows <- rownames(coefs) -->
<!-- # overall_intercept <- fixed_effects[[1]] -->
<!-- coefs <- coefs %>% rename('value' = `(Intercept)`) -->
<!-- for (i in 1:length(rownames(coefs))) { -->
<!--   row.names(coefs)[i] <- paste(rownames(coefs)[i], "(Intercept)", sep = '') -->
<!--   coefs[[1]][i] <- coefs[[1]][i] + fixed_effects[[1]] -->
<!-- } -->
<!-- for (i in 2:8) { -->
<!--   coefs <- coefs %>% add_row(value = fixed_effects[[i]]) -->
<!--   row.names(coefs)[length(rownames(coefs))] <- names(fixed_effects)[i] -->
<!-- } -->
<!-- coefs %>%  -->
<!--   mutate(names = rownames(coefs)) %>%  -->
<!--   pivot_wider(names_from = names, values_from = value) %>%  -->
<!--   kable() -->
<!-- ``` -->

\newpage
# References {.appendix}
::: {#refs}
:::

\newpage
# Appendix: Code {.appendix}
```{r ref.label=knitr::all_labels()}
#| eval: false
#| echo: true
```
